A Meta-Analysis of Music Emotion Recognition Studies

### 2025-03-19 manuscript/manuscript.qmd
updates to manuscript, notes about conversion to pdf

Emotion has been widely discussed since the earliest artificial intelligence (AI) applications to music in the 1950s. Whereas early discourse largely focused on generative composition using computers [@zaripov1969], attention later shifted to creating methods to predict emotion using music's structural cues. Novel techniques for information retrieval emerged in the 1950s and 1960s [@fairthorne1968], inspiring analogous developments for automated music analysis (@kassler1966toward; @mendel1969some). These developments would set the stage for early work in music emotion recognition (MER). Katayose et al. (1988) conducted the first study of this nature, creating an algorithm that associated emotions with analyzed chords to generate descriptions like "there is hopeful mood on chord from 69 to 97 \[*sic*\]." [@katayose_sentiment_1988, p. 1087].

@@

The difficulty in predicting valence reflects a broader challenge in information retrieval. Specifically, relations between low-level predictors from music and text and the perceptual phenomena they model remain poorly understood, reaching a ceiling in prediction accuracy [@celma_foafing_2006]. To address this so-called *semantic gap*, researchers have attempted to identify new feature sets with greater relevance to emotion [@panda2020audio; @chowdhury2021perceived], combine low-, mid-, and high-level features using multimodal data [@celma_foafing_2006], or train neural networks to automatically learn features from audio [@zhang2016br]. Through these approaches, MER researchers attempt to shatter a so-called *glass ceiling* [@downie_music_2008] by establishing central predictors for emotions. To date, however, no study has systematically compared results of these diverse approaches.

@@

Our aim is to evaluate the predictive accuracy of two models of emotional expression in music: (a) models that predict track-specific coordinates in affective circumplex space (valence and arousal), and (b) models that classify discrete emotion categories. We focus on recent to identify key factors such as modeling techniques and features that significantly affect prediction accuracy. To achieve this, we conduct a meta-analysis of journal articles published in the past 10 years. Based on existing literature, we hypothesize that arousal will be predicted with higher accuracy than valence, as valence tends to be more context-dependent and challenging to model. For emotion classification, we expect simple utilitarian emotions (e.g., fear, anger) will be easier to predict than more complex social emotions (e.g., sadness, nostalgia).

@@

|            | Regression                  | Classification            | Total |
|:-----------|:----------------------------|:--------------------------|:------|
| Study N    | 22                          | 12                        | 34    |
| Model N    | 204                         | 86                        | 290   |
| Techniques | Neural Nets: 64             | 21                        | 85    |
| Techniques | Support Vector Machines: 62 | 26                        | 88    |
| Techniques | Linear Methods: 62          | 19                        | 81    |
| Techniques | Tree-based Methods: 14      | 16                        | 30    |
| Techniques | KS, Add. & KNN: 2           | 4                         | 6     |
| Feature N  | Min=3, Md=653, Max=14460    | Min=6, Md=98, Max=8904    | NA    |
| Stimulus N | Min=20, Md=324, Max=2486    | Min=124, Md=300, Max=5192 | NA    |

@@

Since there are many models contained within each of the studies, we will report the results in two parts; We first give an overview of the results for all models, and then we focus on the best performing models of each study. The best performing model is the model within each study with the highest correlation coefficient. This reduction is done to avoid the issue of multiple models from the same study deflating the results as majority of the models included are relative modest baseline or alternative models that do not represent the novelty or content of the article.

@@

Table 2 summarises the results for all models (All) as well as best performing models (Max) for each study for valence. The summary includes the number of models and observations, the correlation coefficient and its 95% confidence interval, the t-value and p-value for the correlation, the heterogeneity statistics $\tau^2$ and $I^2$, calculated through appropriate transformations (Fisher's Z) for the correlation coefficient as part of a random-effects model using `meta` library [@balduzzi2019]. We used Paule-Mandel estimator for between-study heterogeneity [@langan2019comparison] and Knapp-Hartung [@knapp2003improved] adjustments for confidence intervals. In this table we also report two subgroup analyses. One where we have divided the studies according to the number of features they contain (three categories based on quantiles to keep the group size comparable) and into five modelling techniques introduced earlier (Table 1).

@@

To better understand the effects across studies and the nature of the observed heterogeneity, Figure 2 presents (A) a forest and (B) funnel plot of the random-effects model, based on the best-performing models from all studies. In terms of the forest plot, the range of prediction values (correlations) is broad, spanning from 0.13 to 0.92, with all studies except Koh et al. (2023) demonstrating evidence of positive correlations. A mean estimate of 0.67 is achieved by 15 out of the 22 models. While the confidence intervals are generally narrow due to the large sample sizes in each study, there are exceptions, such as smaller sample sizes in @beveridge2018po (n = 20), and in @griffiths2021am (n = 40). The funnel plot in panel B of Figure 2 shows clustering at the top of the plot (studies with low standard error) and no assumed larger diversity in the correlations when the error rates increase. However, there is no clear asymmetry in the plot, verified by non-significant Egger's test ($\beta$ = 5.05, CI95% -0.99-11.09, *t* = 1.64, *p* = 0.112, @eggersmith_1997). Coming back to the mean of valence correlation of 0.669 by all studies and the possible impact of study heterogeneity on this estimation, we also calculated the correlation without the studies that lie outside the 95% CI for pooled effect. This left 12 studies in the data and resulted in the meta-analytical pooled correlation of 0.686 (CI95% 0.635-0.731). In other words, despite the large variation in the correlations and standard errors across the studies, this variation in itself does not seem to be a significant driver behind the overall effect.

@@

@@

Figure 3 presents (A) a forest and (B) funnel plot of the random-effects model of the best-performing models from all studies. Similarly to valence, the range of correlations is also wide for arousal, ranging from 0.35 to 0.95, with all studies demonstrating evidence of positive correlations. A mean estimate of 0.82 or higher is achieved by the majority (14 out of the 22 models). Due to large sample in most studies, the confidence intervals are narrow, although the exceptions ($N < 55$) are clearly visible [@beveridge2018po; @griffiths2021am; @koh2023me; @saizclar2022pr; @wang2021ac]. The funnel plot in panel B illustrates the limited range of the correlations and show heavy clustering at the top of the plot (studies with low standard error). There is no clear asymmetry in the plot, verified by non-significant Egger's test reported above.

@@

@@

### Model success across concepts

To assess how the use of different model types affected performance, we prepared heatmap visualizations depicting differences in success across feature *n* categories. We collapsed SVM and Tree-Based categories due to their low representation in the model summary. Figure 5 summarizes differences in success across categories (a), as well as the algorithms in each model class (b). Oveall, studies using smaller feature sets tend to perform best, whereas the best model type largely depends on the nature of the prediction task.

![Heatmaps summarizing models. 5(a) depicts overall success across tasks according to model type and feature *n* summaries. Text labels summarize the number of models belonging to each subgroup, whereas the numbers in parentheses summarize the median number of stimuli used. 5(b) summarizes the algorithms pertaining to each class, colour-coded by frequency. Grey rectangles are visual aids aligning with *x*-axis groupings in 5(a) (white: LM; light grey: NN; dark grey: other).](../manuscript/model-summary-fig.png "Heatmap depicting success across tasks according to Feature N category and model class. Labels indicate the number of studies and the median number of stimuli is shown in parentheses."){width="668"}

>>>>

Emotion has been widely discussed since the earliest artificial intelligence (AI) applications to music in the 1950s. Whereas early discourse largely focused on generative composition using computers [@zaripov1969], attention later shifted to creating methods to predict emotion using music's structural cues. Novel techniques for information retrieval emerged in the 1950s and 1960s [@fairthorne1968], inspiring analogous developments for automated music analysis [@kassler1966toward; @mendel1969some]. These developments would set the stage for early work in music emotion recognition (MER). @katayose_sentiment_1988 conducted the first study of this nature, creating an algorithm that associated emotions with analyzed chords to generate descriptions like "there is hopeful mood on chord from 69 to 97 \[*sic*\]." [@katayose_sentiment_1988, p. 1087].

@@

The difficulty in predicting valence reflects a broader challenge in information retrieval. Specifically, relations between low-level predictors from music and text and the perceptual phenomena they model remain poorly understood, reaching a ceiling in prediction accuracy [@celma_foafing_2006]. To address this so-called *semantic gap*, researchers have attempted to identify new feature sets with greater relevance to emotion [@panda2020audio; @chowdhury2021perceived], combine low-, mid-, and high-level features using multimodal data [@celma_foafing_2006], or train neural networks to automatically learn features from audio [@zhang2016br;@alvarez2023ri;@agarwal2021an]. Through these approaches, MER researchers attempt to shatter a so-called *glass ceiling* [@downie_music_2008] by establishing central predictors for emotions. To date, however, no study has systematically compared results of these diverse approaches.

<!-- added more recent references to make the neural network mention more up to date, TE 19/3/2025 -->

@@

Our aim is to evaluate the predictive accuracy of two models of emotional expression in music: (a) models that predict track-specific coordinates in affective circumplex space (valence and arousal), and (b) models that classify discrete emotion categories. We focus on recent studies to identify the overall success rate in MER tasks and the key factors such as modeling techniques and features that contribute to the prediction accuracy of the models. To achieve this, we conduct a meta-analysis of journal articles published in the past 10 years. Based on existing literature, we hypothesize that arousal will be predicted with higher accuracy than valence, as valence tends to be more context-dependent and challenging to model. For emotion classification, we expect simple utilitarian emotions (e.g., fear, anger) will be easier to predict than more complex social emotions (e.g., sadness, nostalgia).

<!-- DISCUSS: we have not address this last thought at all, TE 19/3/2025 -->

@@

|            | Regression                  | Classification            | Total |
|:-----------|:----------------------------|:--------------------------|:------|
| Study N    | 22                          | 12                        | 34    |
| Model N    | 204                         | 86                        | 290   |
| Techniques | Neural Nets (NN): 64             | 21                        | 85    |
| Techniques | Support Vector Machines (SVM): 62 | 26                        | 88    |
| Techniques | Linear Methods (LM): 62          | 19                        | 81    |
| Techniques | Tree-based Methods (TM): 14      | 16                        | 30    |
| Techniques | KS, Add. & KNN: 2           | 4                         | 6     |
| Feature N  | Min=3, Md=653, Max=14460    | Min=6, Md=98, Max=8904    | NA    |
| Stimulus N | Min=20, Md=324, Max=2486    | Min=124, Md=300, Max=5192 | NA    |

@@

Since there are many models contained within each of the studies, we will report the results in two parts; We first give an overview of the results for all models, and then we focus on the best performing models of each study. The best performing model is the model within each study with the highest correlation coefficient. This reduction is done to avoid the issue of multiple models from the same study deflating the results as majority of the models included are relative modest baseline or alternative models that do not represent the novelty or content of the article. We also provide a summary of the results with all models included in addition to the chosen strategy, where the best model of each study is considered.

@@

Table 2 summarises the results for all models (All) as well as best performing models (Max) for each study for valence. The summary includes the number of models and observations, the correlation coefficient and its 95% confidence interval, the t-value and p-value for the correlation, the heterogeneity statistics $\tau^2$ and $I^2$, calculated through appropriate transformations (Fisher's Z) for the correlation coefficient as part of a random-effects model using `meta` library [@balduzzi2019]. We used Paule-Mandel estimator for between-study heterogeneity [@langan2019comparison] and Knapp-Hartung [@knapp2003improved] adjustments for confidence intervals. In this table we also report two subgroup analyses. One where we have divided the studies according to the number of features they contain (three categories based on quantiles to keep the group size comparable) and into four modelling techniques introduced earlier (Table 1).

<!-- DISCUSS: I have changed the modelling techniques into FOUR categories, TE 19/3/2025 -->

@@

To better understand the effects across studies and the nature of the observed heterogeneity, Figure 2 presents a forest of the random-effects model, based on the best-performing models from all studies. In terms of the forest plot, the range of prediction values (correlations) is broad, spanning from 0.13 to 0.92, with all studies except Koh et al. (2023) demonstrating evidence of positive correlations. A mean estimate of 0.67 is achieved by 15 out of the 22 models. While the confidence intervals are generally narrow due to the large sample sizes in each study, there are exceptions, such as smaller sample sizes in @beveridge2018po (n = 20), and in @griffiths2021am (n = 40). If we explore the asymmetry of the model prediction rate across standard error, we do not observe particular asymmetries that would indicate particular bias in the reported studies. This is verified by non-significant Egger's test ($\beta$ = 5.05, CI95% -0.99-11.09, *t* = 1.64, *p* = 0.112, @eggersmith_1997).

Coming back to the mean of valence correlation of 0.669 by all studies and the possible impact of study heterogeneity on this estimation, we also calculated the correlation without the studies that lie outside the 95% CI for pooled effect. This left 12 studies in the data and resulted in the meta-analytical pooled correlation of 0.686 (CI95% 0.635-0.731). In other words, despite the large variation in the correlations and standard errors across the studies, this variation in itself does not seem to be a significant driver behind the overall effect.

<!-- numbers in the text updated up to here, CA 20/2/2025 -->
<!-- I've removed the reference to the funnel plot, TE 19/3/2025 -->

@@

<!-- numbers in the text updated up to here, CA 20/2/2025 -->
<!-- DISCUSS: we don't report journal type of explain/predict differences, although we do some of these in arousal, TE 19/3/2025 -->

@@

Figure 3 presents a forest plot of the random-effects model of the best-performing models from all studies. Similarly to valence, the range of correlations is also wide for arousal, ranging from 0.35 to 0.95, with all studies demonstrating evidence of positive correlations. A mean estimate of 0.81 or higher is achieved by the majority (14 out of the 22 models). Due to large sample in most studies, the confidence intervals are narrow, although the exceptions ($N < 55$) are clearly visible [@beveridge2018po; @griffiths2021am; @koh2023me; @saizclar2022pr; @wang2021ac].

<!-- Removed reference to the funnel plot, TE 19/3/2025 -->

@@

<! -- DISCUSS: 4 digits is too much, let's make this consistent, 3 in all places?, TE 19/3/2025 -->

@@

### Model success across concepts, model types and feature counts

<!-- Changed the header slightly, TE 19/3/2025-->

To assess how the use of different model types affected performance, we prepared heatmap visualizations depicting differences in success across feature *n* categories. We collapsed SVM and Tree-Based categories due to their low representation in the model summary. Figure 5 summarizes differences in success (a) across categories, as well as (b) the algorithms in each model class. Overall, studies using smaller feature sets tend to perform best, whereas the best model type largely depends on the nature of the prediction task.

![Heatmaps summarizing models. Panel a depicts overall success across tasks according to model type and feature *n* summaries. Text labels summarize the number of models belonging to each subgroup, whereas the numbers in parentheses summarize the median number of stimuli used. Panel b summarizes the algorithms pertaining to each class, colour-coded by frequency. Grey rectangles are visual aids aligning with *x*-axis groupings in panel a (white: LM; light grey: NN; dark grey: other).](../manuscript/model-summary-fig.png "Heatmap depicting success across tasks according to Feature N category and model class. Labels indicate the number of studies and the median number of stimuli is shown in parentheses."){width="668"}
